Large-scale Omics Experiments are not Conducive to Causal Inference

### 2015-09-07 write_up.Rmd
incorporated Olgas comments into Karens

---
title: "Distilling Causality from Correlation in Large-scale Omics Datasets"
author: "Robert Ness"
date: "September 3, 2015"
output: html_document
---

Our goal in molecular cell biology is causal inference -- meaning uncovering the regulatory relationships between components of the cell that determine phenotype and disease states.  Generally observing statistical association between such components does not demonstrate causality.  However, we can safely use statistical association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.

High-throughput omics technologies enable us to conduct large-scale experiments, where we simultaneously measure multiple cellular components in a sample, and in many cases acquire a large amount of data for each component.  However, despite generating much more data, these large-scale experiments make the task of causal inference much more difficult.  Large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present.  We can address this problem by improving our experimental design.

## How we infer causality from omics experiments

The MAPK signaling pathway proteins Mek and Erk have a cause-and-effect relationship; Mek has a causal impact on Erk. The causal mechanism is phosphorylation, Mek phosphorylates Erk.

Assuming this causal relation were not known, the question of whether Mek has a causal effect on Erk is evaluated with a designed experiment, where Mek is the treatment and Erk is the response.  Intervention and randomization are key parts of the experimental design.  Intervention means we target the potential cause, Mek, and fix it at different levels, for example 'high activity' and 'baseline activity'.  Randomization refers to random assignment of samples or subjects to each of these levels, helping to alleviate bias introduced by unseen factors.  The experiment generates measurements that will show a strong statistical association between Mek and Erk.  Intervention and randomization enable us to interpret that statistical association as evidence for Mek's causal influence on Erk.

With omics technologies, instead of measuring one response, we simultaneous measure many.  Instead of one statistical association to evaluate, the measurements contain many pairwise associations between each of the measured features.

As with the simple one-response experimental design, we seek to use these associations to evaluate the presence of causal relations.  Saying two objects have a statistical association is the same as saying they are not statistically independent.  However, they can still be conditionally independent, meaning the association disappears conditional on knowing what all the other components in the system are doing. Figure 1 illustrates this principle with the MAPK proteins.

<figure>
  <img src="http://i.imgur.com/Ft2exsrl.png" alt="Mapk" width="304" height="228">
  <figcaption>Fig1. In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk.
  Top; the lines represent associate - the activity of the kinases on this pathway are all correlated. Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk. They are conditionally independent.</figcaption>
</figure>

We use algorithms borrowed from the causal Bayesian network modeling domain to do causal inference in the context of an experiment with multiple intercorrelated response variables.  These  algorithms' approaches vary but they essentially do two things; (1) find cases of conditional independence and reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use the experiment's intervention information to evaluate those conditional dependencies as evidence for potential causal relationtions.

The challenge with large datasets is that by measuring more things, we increase the number of spurious statistical associations, and thus the number of spurious cases of conditional dependence, and thus the number of incorrect conclusions of the presence of a causal relation.  To illustrate, we ran a simulation where first simulate a 20 features dataset each with a 100 Gaussian random measurements, then increase the number of features to 500.  In each case the features are completely independent.  This means if we quantify association using Pearson correlation, any correlation we find will be completely spurious.  We record the maximum correlation between the 20 feature set and the 500 feature set. We repeat this simulation 500 times.

@@

The increased incidence of spurious correlation means increased false positives in detecting conditional dependence relationships.  To illustrate, we repeat the previous simulation, except now expanding from 20 to only 100 features. In each of the 500 instances, we apply a search algorithm that iteratively performs conditional independence tests, returning a count of detected conditional dependence relationships.  As before, since we simulated independent features, any conditional dependence relationship reported by this algorithm is false.

@@

Lastly, measuring more things also adversely effects the selection of interventions in the design.  The more components of the cell you measure, the more interventions you need to target those components.  This, of course, increases the cost of the experiment.  Under constrained resources, we can use a limited set of interventions that will partially elucidate causality in the system.  We use prior biological knowledge to prioritize what to target with that limited set of interventions.  For example, if we know what proteins a candidate drug is likely to impact, then interventions should target those proteins and those known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.

## Use of prior biological knowledge to address the spurious association problem

Similarly, we can use to prior biological knowledge to address the spurious association problem and improve detection of conditional dependence relationships.  Specifically, this approach uses pathway databases such as KEGG to answer the questions;

1) What causal relations are we sure exist? 
2) What causal relations are we sure do not exist? 
3) What causal relations do we expect to exist/not to exist?

Answering these questions enables us to improve the search for conditional dependence relations, minimizing false positives do to spurious associations.

>>>>

---
title: "Large-scale Omics Experiments are not Conducive to Causal Inference"
date: "September 3, 2015"
output: html_document
---

Causal inference – the task of uncovering regulatory relationships between components of biomolecular pathways and networks - is a primary goal of omics datasets in molecular biology studies. Generally, observing statistical association between network components does not demonstrate causality. However, we can safely use association observed in measurements generated by a sound experimental design as evidence for the presence of a causal relation.

High-throughput omics technologies enable large-scale experiments, with simultaneous measurements of multiple cellular components in a sample. The huge amount of data generated paradoxically make the task of causal inference much more difficult because, when many examining pairwise associations between measured features, some will appear significant just by chance. In other words, large-scale experiments are more prone to hiding the true signal and generating spurious associations, leading to increased false positive conclusions that causal relationships are present. This problem can be addressed by improving experimental design and refining the biological question. Below, we describe the process of elucidating statistical associations from data and suggest practical approaches for analysis of large scale datasets.

## Increasing the number of measured features decreases the ability to infer causality

In MAPK signaling, Raf causally affects active Mek levels (via phosphorylation), while Mek causally affects Erk. Imagine these relationships were unknown: could it be detected from measurements of these phosphoproteins?  If we took an arbitrary biological sample and simultaneously quantified Raf, Mek, and Erk, we'd find Raf-Mek, Mek-Erk, and Raf-Erk correlations, due their direct and indirect causal relationships.

To get from correlation to causality, we need a controlled experiment.  Based on the Mek-Erk correlation, we can design an experiment based on the hypothesis that Mek has a causal influence on Erk.  In the experiment, we'd *intervene* in Mek, meaning we'd artificially control Mek at different levels of phospho-activity (for example with a small molecule treatment that inhibits Mek), then randomly assign cell lines to each of these levels.  In doing so we'd control for confounding factors and free ourselves to interpret the correlation between Mek levels and Erk measurements as evidence for the presence of causal influence of Mek on Erk.  Indeed, if our intervention targeted Erk and we observed Mek, the measured correlation would be very weak since the Erk->Mek causal relationship doesn't exist.

Now, suppose you have access to a technology that can simultaneously quantify several thousands of phosphoproteins, including Raf, Mek and Erk.  As is typical with such large-scale experiments, you only have a very small number of interventions, and maybe 3 replicates.  Even if those interventions targeted Raf, Mek, or Erk, there would be so much spurious correlation between these phosphoproteins and all the others you mentioned that causal inference would be infeasible.

To explain why, we need to introduce some terminology.  A statistical association (such as Pearson correlation) between two measured entities indicates that they are dependent. Sometimes we choose to assess the dependence in the context of already knowing what other key components of the system are doing.  In this case the dependence is called *conditional dependence*, because we evaluate the association conditional on the states of other components in the system.

We can see how this applies here by examining the dependence between Raf and Erk.  Figure 1, top panel, depicts the correlation between Raf, Mek, and Erk as black lines, causal relationships are depicted as blue arrows.  If we condition on Mek, then we assume that we know what Mek is doing.  In all the measurements where Mek is high, can the dependence between Raf and Erk still be detected?  Since Mek is directly upstream of Erk, knowing Mek tells us all there is to know about Erk, and knowing Raf tells us nothing.  Because of that, the Raf-Erk dependence, once we condition on Mek, will be reduced or eliminated, thus indicating that the Raf-Erk edge can be removed.  Using this approach, it is possible to keep only the conditional dependence edges as depicted by the black lines in Figure 1, bottom panel.  Conditional dependence is important because, under certain assumptions, we can move from conditional dependence to causal influence by applying interventions.

<figure>
  <img src="http://i.imgur.com/Ft2exsrl.png" alt="Mapk" width="304" height="228">
  <figcaption>
  Fig1. In the MAPK signaling pathway, Raf regulates Mek, which regulates Erk. Top; the lines represent association - the activity of the kinases on this pathway are all correlated. Bottom: the lines represent conditional dependence, there is no conditional dependence edge from Raf to Erk because given you know Mek, knowing Raf tells you nothing about Erk. They are conditionally independent.
  </figcaption>
</figure>

Computational methods that do causal inference in the context of an experiment that simultaneously measures multiple intercorrelated response variables all essentially have the same two step process; (1) reduce the dense set of pairwise associations to a sparse set of conditional dependencies, (2) use information about the interventions included in the experiment to evaluate those conditional dependencies as evidence for potential causal relations.

The problem of doing causal inference with large-scale experiments is two-fold.  Firstly, sample size is typically far too small for the large number of features we measure.  With the large-scale experiment described above, it is clear that viewing a scatter-plot between a pair of features that has only 3 points is uninformative -- even if the points correlated perfectly, 3 points is not enough to conclude anything about the relationship with any confidence.  But even with hundreds of samples, we are still plagued by spurious associations when the number of features is so large, since more features means greater chance associations will occur at random.

To illustrate, we ran a simulation where we first simulate a 20 features dataset each with a 100 Gaussian random measurements, then increase the number of features to 500. In both cases the features are completely independent. This means if we quantify association using Pearson correlation, any correlation we find will be completely spurious. We record the maximum correlation between the 20 feature set and the 500 feature set. We repeat this simulation 500 times.

@@

The increased incidence of spurious correlation means increased false positives in detecting conditional dependence relationships. To illustrate, we repeat the previous simulation, except now expanding from 20 to only 100 features. In each of the 500 instances, we apply a search algorithm that iteratively performs conditional independence tests, returning a count of detected conditional dependence relationships. As before, since we simulated independent features, any conditional dependence relationship reported by this algorithm is a false positive, i.e. it doesn't actually exist in the mechanism that generated the data.

@@

As you increase the number of features you measure, the number of false positive detection of conditional dependence explodes.  What this means is that the computational methods for causal inference will fail for the typical large-scale experiment, because when their input data contains a large amount of features relative to sample size, they cannot reliably detect the conditional dependence relationships they need to infer causality.

The second problem is that the aforementioned computational methods require interventions to to infer causality.  To infer the causal relationships between many measured features, you need many interventions targeting those many features.  Thousands of interventions targeting thousands of features is infeasible.

## Experimentalists can apply heuristics for causal inference in large-scale omics data

Fortunately, there are ways to address this problem, these *Causality Inference in Omics* heuristics are listed below:

1) Measure many samples.  If feasible, high throughput measurements that measure many samples (not just many features) provide the statistical power to tell true associations from spurious associations.  This intuition motivates the use of single cell data for causal modeling, where having many thousands of cells per sample provides statistical power.

2) Refine the problem, thus limiting the number of features.  If the biological system in question is sufficiently well-studied, it might be possible to ask specific questions of the data, such as ascertaining the presence or absence of a particular edge or pathway, or assessing connections to a particular protein.  The more specific the question, the less data is needed overall to make solid statistical inferences.

3) Use prior knowledge, not just from experts but also from noisier sources.  Based on information in  pathway databases such as KEGG, answer the questions;

i) What causal relations are we sure exist? 
ii) What causal relations are we sure do not exist? 
iii) What causal relations do we expect to exist/not to exist?

Such questions can be answered with a variety of biological knowledge.  For example, we know more about causal relations between phosphoproteins known to cluster together spatially within the cell, and be more interested in learning about signaling mechanisms that cross compartments inside the cell. By answering these questions, we can bias the application of computational causal inference questions such that they focus less on conditional dependence and causal relations we know to exist and more towards the novel causal information hidden in the data.

4) Employ interventions selectively.  Some interventions are 'fat-handed', in that they perturb many features in the system, providing more causal inference bang for your intervention buck.  At the same time, targeted interventions are more useful for gaining mechanistic insight into the biological question under study.  For example, if proteins a candidate drug is likely to impact are known, then interventions could target those proteins and other proteins known a priori to interact with them, in order to acquire data that elucidates the mechanism of action for that drug.

The Causality in Omics tool list above provides impactful approaches that can drastically improve causal inference from omics datasets, by constraining the inference task, rendering it much smaller, and thus allowing for accurate statistical inferences.  For instance, the task of assessing which of all the possible KEGG pathways is present in my dataset is a far smaller (thus easier) task statistically speaking than the task of assessing which of all possible combinations of my measured features might form a biological pathway.

How should the tools listed be used?  They are most powerful when used in combination, and in fact the lines between them are somewhat arbitrary and frequently blurred.  For instance, tool #4 calls for use of interventions, but this task itself is complicated by large scale data, as the number of possible perturbations grows and experiments quickly become infeasible. Tool 3, prior biological knowledge, can be used to prioritize what to target with that limited set of interventions.
